Fail Faster

### 2025-03-20 paper/paper.tex
dashes

The precise details of how the SplitMix algorithm works \cn are unimportant for the
present paper, but the critical component is that all of its operations are are
defined in terms of arithmetic bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory.
This has a significant impact on the performance of generators. A single call to
one of the \bq library functions --- like generating an integer uniformly in a range --- may call the \texttt{splittable\_random}
algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
times \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently collect in OCaml,
\footnote{The OCaml GC is a generational collector \cn, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}, we will see in Section~\ref{section:eval} that this can have
a large performance impact on some generators that spend most of their time sampling data.
To circumvent this allocation and provide an equivalent version of \texttt{splittable\_random}, we
reimplement SplitMix in C, and call out to it with the OCaml FFI. The C version of the library uses
proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and C code.
Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
discussed in the previous sections.

@@

For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
using a variety of \textit{strategies}, which vary in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a custom, correct-by-construction generator of
STLC terms, whose behavior is the most complex. 
For evaluating \scalaname, we implement a subset of the afformentioned strategies  ---
only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.

@@

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance gain
--- purely from staging --- over ScalaCheck than \name does over \bq. This discrepancy arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}. Each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging allows us to eliminate this
overhead, resulting in substantial speedups in aggregate.

>>>>

The precise details of how the SplitMix algorithm works \cn are unimportant for the
present paper, but the critical component is that all of its operations are are
defined in terms of arithmetic bitwise operations on 64-bit integers. In OCaml,
because of details related to the garbage collector, the 64-bit integer type \texttt{int64} is represented
at run time as a \emph{pointer} to an unscanned block of memory containing (among other things)
a 64-bit integer \cn. This means that all operations that return an \texttt{int64} must allocate this block of memory.
This has a significant impact on the performance of generators. A single call to
one of the \bq library functions---like generating an integer uniformly in a range---may call the \texttt{splittable\_random}
algorithm multiple times. Each sample from \texttt{splittable\_random} allocates 9
times \jwc{this is a call to \texttt{next\_int64}}, and each allocation brings us closer
to the next garbage collection pause. While small allocations like these are \emph{very} fast to perform and subsequently collect in OCaml,
\footnote{The OCaml GC is a generational collector \cn, and since these allocations are small and mostly very short lived, they will all be minor allocations, never to be promoted.}, we will see in Section~\ref{section:eval} that this can have
a large performance impact on some generators that spend most of their time sampling data.
To circumvent this allocation and provide an equivalent version of \texttt{splittable\_random}, we
reimplement SplitMix in C, and call out to it with the OCaml FFI. The C version of the library uses
proper \texttt{int64\_t} arithmetic, only boxing and unboxing integers at the call boundaries between OCaml and C code.
Ideally in the future, one would not need to call out to C for this: the Jane Street bleeding-edge OCaml compiler has support
for unboxed types \cn, which (among other things) would let us implement a version of SplitMix that does not allocate, directly
in OCaml. Unfortunately, the Jane Street branch of the compiler is incompatible with MetaOCaml, which we use to implement the metaprogramming
discussed in the previous sections.

@@

For \camlname, our test cases consist of generators for boolean lists, binary search trees (BST), and simply
typed lambda calculus (STLC) terms. We implement generators for these benchmarks
using a variety of \textit{strategies}, which vary in structure and
sophistication. These include type-derived generators for BSTs and STLC terms,
using the approach outlined in \secref{subsection:type-derived}; two custom BST
generators, one that repeatedly inserts values into an initially empty tree and
one that materializes the tree in a single pass by generating keys, values, and
subtrees at each step; a generator of boolean lists that operates similarly to
the single-pass BST strategy; and a custom, correct-by-construction generator of
STLC terms, whose behavior is the most complex. 
For evaluating \scalaname, we implement a subset of the afformentioned 
strategies---only the boolean list generator, single-pass BST generator, and well-typed STLC term generator.

@@

As shown in \figref{fig:scala}, \scalaname achieves an even greater performance 
gain---purely from staging---over ScalaCheck than \name does over \bq. This discrepancy arises from
ScalaCheck's representation of generators as functions of type \texttt{size ->
seed -> Option[A]}. Each generator combinator must construct and then
pattern-match on these \texttt{Option} values, introducing significant boxing
and unboxing overhead at each step.  Staging allows us to eliminate this
overhead, resulting in substantial speedups in aggregate.
